Alex's Notes and Text Fragments for the JOSS Manuscript

### 2021-05-02 paper/paper.Rmd
adjust caching settings
reorder feature space section

add content to spatial tiling section

---
documentclass: jss
author:
  - name: Patrick Schratz
    affiliation: 'Friedrich-Schiller-University Jena \AND'
    # affiliaton2: 'Ludwig-Maximilians-University Munich'
    # use this syntax to add text on several lines
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{patrick.schratz@uni-jena.de}
  - name: Marc Becker
    affiliation: 'Ludwig-Maximilians-University Munich'
    address: |
      Department of Statistics
      Statistical Learning and Data Science group
    # To add another line, use \AND at the end of the previous one as above
  - name: Michel Lang
    affiliation: TU Dortmund University
    address: |
      | Faculty of Statistics
  - name: Alexander Brenning
    affiliation: 'Friedrich Schiller University Jena'
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{alexander.brenning@uni-jena.de}
title:
  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for \\pkg{mlr3}"
  # If you use tex in the formatted title, also supply version without
  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for mlr3"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com) [@mlr3].
  It adds support for spatiotemporal resampling methods and respective visualization methods.
  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatio-temporal autocorrelation (STAC)) in datasets.
  These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@sperrorest].
  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017]

@@

```{r setup-2, echo=FALSE, purl=TRUE}
# for knitr spint later
knitr::opts_chunk$set(fig.path = "./paper/figs/")
```

@@

```{r create task, cache=F}
library(mlr3)
library(mlr3spatiotempcv)

@@

```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="ht", fig.height=6, fig.width=8, cache=TRUE}
resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
resampling_sp$instantiate(task)

@@

```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="ht", fig.height=6, fig.width=8, cache=T}
resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
resampling_nsp$instantiate(task)

@@

```{r buffer, fig.cap="Visualization of 'Spatial buffering' method.", out.width="50%", cache=T}
resampling_buffer = rsmp("spcv_buffer", theRange = 1000)
resampling_buffer$instantiate(task)

plot(resampling_buffer,
  size = 0.8, task = task, fold_id = 1, crs = 32717) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r block-random, cache = TRUE, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="50%", fig.pos="ht"}
resampling_block_random = rsmp("spcv_block", range = 1000, folds = 5)

plot(resampling_block_random,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r block-systematic, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="50%", fig.pos="ht"}
resampling_block_systematic = rsmp("spcv_block",
  range = 1000, folds = 5,
  selection = "systematic"
)

plot(resampling_block_systematic,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r block-checkerboard, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="40%", fig.pos="h!"}
resampling_block_checkerboard = rsmp("spcv_block",
  range = 1000, folds = 5,
  selection = "checkerboard")

plot(resampling_block_checkerboard,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r block-rows-and-cols, cache=TRUE, fig.cap="Spatial blocking with arguments `rows` and `cols`.", out.width="40%", fig.pos="ht"}
resampling_block_rows_cols = rsmp("spcv_block",
  rows = 6, cols = 3, folds = 5
)

plot(resampling_block_rows_cols,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

@@

```{r coords, fig.cap="Method 'Spatial CV' showing a k-means clustering approach based on coordinates.", cache=TRUE, out.width="50%", fig.pos="ht"}
resampling_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)

plot(resampling_coords,
  size = 0.8, fold_id = 1, task = task, crs = 32717) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

### Feature space clustering (blockCV)

```{r env-1, fig.cap="Environmental blocking using a single predictor.", cache=F, out.width="50%", fig.pos="ht"}
resampling_env = rsmp("spcv_env",
  features = "distdeforest", folds = 5)

plot(resampling_env,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

```{r env-2, fig.cap="Environmental blocking using multiple predictors.", cache=T, out.width="50%", fig.pos="ht"}
resampling_env_multi = rsmp("spcv_env",
  features = c("distdeforest", "slope"), folds = 5)

plot(resampling_env_multi,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r, results='hide', cache=TRUE}
data = cookfarm_sample
data$Date = rep(c(
  "2020-01-01", "2020-02-01", "2020-03-01", "2020-04-01",
  "2020-05-01"), times = 1, each = 100)
b = mlr3::as_data_backend(data)
b$hash = "_mlr3_tasks_cookfarm_"
task_spt = TaskRegrST$new(
  id = "cookfarm", b, target = "PHIHOX",
  extra_args = list(
    coordinate_names = c("x", "y"), coords_as_features = FALSE,
    crs = 26911)
)

@@

plot(resampling_cstf_time,
  fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE
)
```

@@

```{r cache=TRUE, results='hide'}
resampling_cstf_loc = rsmp("sptcv_cstf", folds = 5, space_var = "SOURCEID")
resampling_cstf_loc$instantiate(task_spt)

plot(resampling_cstf_loc,
  fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE)
```

@@

```{r cache=TRUE, results='hide'}
resampling_cstf_time_loc = rsmp("sptcv_cstf",
  folds = 5,
  space_var = "SOURCEID", time_var = "Date")
resampling_cstf_time_loc$instantiate(task_spt)

plot(resampling_cstf_time_loc,
  fold_id = 4, task = task_spt, crs = 32717, plot3D = TRUE,
  show_omitted = TRUE)
```

```{r llto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-And-Time-Out' (LLTO). The grey points show the points omitted by this method.", cache=TRUE, fig.pos="ht"}
knitr::include_graphics("llto.png")
```

@@

@@

```{r prepare, cache=TRUE}
library("mlr3")
library("mlr3spatiotempcv")
library("mlr3learners")

@@

```{r rr-nspcv, cache=TRUE, cache.lazy = FALSE}
resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
resampling_nsp$instantiate(task)
rr_nsp = resample(
  task = task, learner = learner,
  resampling = resampling_nsp
)

@@

```{r rr-spcv, cache=TRUE, cache.lazy = FALSE}
resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
resampling_sp$instantiate(task)
rr_sp = resample(
  task = task, learner = learner,
  resampling = resampling_sp
)

>>>>

---
documentclass: jss
author:
  - name: Patrick Schratz
    affiliation: 'Friedrich-Schiller-University Jena'
    # affiliaton2: 'Ludwig-Maximilians-University Munich'
    # use this syntax to add text on several lines
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{patrick.schratz@uni-jena.de}
  - name: Marc Becker
    affiliation: 'Ludwig-Maximilians-University Munich \AND'
    address: |
      Department of Statistics
      Statistical Learning and Data Science group
    # To add another line, use \AND at the end of the previous one as above
  - name: Michel Lang
    affiliation: TU Dortmund University
    address: |
      | Faculty of Statistics
  - name: Alexander Brenning
    affiliation: 'Friedrich Schiller University Jena'
    address: |
      | Department of Geography
      | Geographic Information Science group
    email: \email{alexander.brenning@uni-jena.de}
title:
  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for \\pkg{mlr3}"
  # If you use tex in the formatted title, also supply version without
  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for mlr3"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com) [@mlr3].
  It adds support for spatiotemporal resampling methods and respective visualization methods.
  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatio-temporal autocorrelation (STAC)) in datasets.
  These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@sperrorest].
  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017]

@@

```{r setup-2, echo=FALSE, purl=T}
# for knitr spint later
knitr::opts_chunk$set(fig.path = "./paper/figs/",
  cache = F,
  cache.lazy = TRUE)
```

@@

```{r create task}
library(mlr3)
library(mlr3spatiotempcv)

@@

```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="ht", fig.height=6, fig.width=8}
resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
resampling_sp$instantiate(task)

@@

```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="ht", fig.height=6, fig.width=8}
resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
resampling_nsp$instantiate(task)

@@

```{r buffer, fig.cap="Visualization of 'Spatial buffering' method.", out.width="50%"}
resampling_buffer = rsmp("spcv_buffer", theRange = 1000)
resampling_buffer$instantiate(task)

autoplot(resampling_buffer,
  size = 0.8, task = task, fold_id = 1, crs = 32717) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r block-random, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="50%", fig.pos="ht"}
resampling_block_random = rsmp("spcv_block", range = 1000, folds = 5)

autoplot(resampling_block_random,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r block-systematic, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="50%", fig.pos="ht"}
resampling_block_systematic = rsmp("spcv_block",
  range = 1000, folds = 5,
  selection = "systematic"
)

autoplot(resampling_block_systematic,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r block-checkerboard, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="40%", fig.pos="h!"}
resampling_block_checkerboard = rsmp("spcv_block",
  range = 1000, folds = 5,
  selection = "checkerboard")

autoplot(resampling_block_checkerboard,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r block-rows-and-cols, fig.cap="Spatial blocking with arguments `rows` and `cols`.", out.width="40%", fig.pos="ht"}
resampling_block_rows_cols = rsmp("spcv_block",
  rows = 6, cols = 3, folds = 5
)

autoplot(resampling_block_rows_cols,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

Method `partition_tiles()` from package \pkg{sperrorest} also uses rectangular blocks to partition the study area.
In contrast to "Spatial Blocking" from \pkg{blockCV} these blocks can be rotated (via argument `rotation`) and a minimum number of observations per block can be set (argument `min_n`).
A predefined row/column grid is required via either argument `dsplit` or `nsplit` and no square blocks can be used.
Small tiles can be merged with adjacent blocks via argument `reassign` or optionally be ignored (`reassign = FALSE`).
The number of tiles set by the user equals the number of folds and blocks cannot be grouped into folds as in "Spatial Blocking".
Also despite setting 12 blocks via `nsplit = c(3L, 4L)` in the following example, the final partition count is eleven because the bottom-left part of the study area does not contain observations.

```{r tile, fig.cap="Spatial tiling with arguments `rows` and `cols`.", out.width="40%", fig.pos="ht"}
resampling_tiles = rsmp("spcv_tiles",
  nsplit = c(3L, 4L)
)

autoplot(resampling_tiles,
  size = 0.8, fold_id = 1, task = task, crs = 32717) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r coords, fig.cap="Method 'Spatial CV' showing a k-means clustering approach based on coordinates.", out.width="50%", fig.pos="ht"}
resampling_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)

autoplot(resampling_coords,
  size = 0.8, fold_id = 1, task = task, crs = 32717) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r, results='hide'}
data = cookfarm_sample
data$Date = rep(c(
  "2020-01-01", "2020-02-01", "2020-03-01", "2020-04-01",
  "2020-05-01"), times = 1, each = 100)
b = mlr3::as_data_backend(data)
b$hash = "_mlr3_tasks_cookfarm_"
task_spt = TaskRegrST$new(
  id = "cookfarm", b, target = "PHIHOX",
  extra_args = list(
    coordinate_names = c("x", "y"), coords_as_features = FALSE,
    crs = 26911)
)

@@

autoplot(resampling_cstf_time,
  fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE
)
```

@@

```{r results='hide'}
resampling_cstf_loc = rsmp("sptcv_cstf", folds = 5, space_var = "SOURCEID")
resampling_cstf_loc$instantiate(task_spt)

autoplot(resampling_cstf_loc,
  fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE)
```

@@

```{r results='hide'}
resampling_cstf_time_loc = rsmp("sptcv_cstf",
  folds = 5,
  space_var = "SOURCEID", time_var = "Date")
resampling_cstf_time_loc$instantiate(task_spt)

autoplot(resampling_cstf_time_loc,
  fold_id = 4, task = task_spt, crs = 32717, plot3D = TRUE,
  show_omitted = TRUE)
```

```{r llto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-And-Time-Out' (LLTO). The grey points show the points omitted by this method.", fig.pos="ht"}
knitr::include_graphics("llto.png")
```

@@

## Feature space clustering

```{r env-1, fig.cap="Environmental blocking using a single predictor.", out.width="50%", fig.pos="ht"}
resampling_env = rsmp("spcv_env",
  features = "distdeforest", folds = 5)

autoplot(resampling_env,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

```{r env-2, fig.cap="Environmental blocking using multiple predictors.", out.width="50%", fig.pos="ht"}
resampling_env_multi = rsmp("spcv_env",
  features = c("distdeforest", "slope"), folds = 5)

autoplot(resampling_env_multi,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r prepare}
library("mlr3")
library("mlr3spatiotempcv")
library("mlr3learners")

@@

```{r rr-nspcv, cache.lazy = FALSE}
resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
resampling_nsp$instantiate(task)
rr_nsp = resample(
  task = task, learner = learner,
  resampling = resampling_nsp
)

@@

```{r rr-spcv, cache.lazy = FALSE}
resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
resampling_sp$instantiate(task)
rr_sp = resample(
  task = task, learner = learner,
  resampling = resampling_sp
)
